Parametricity Consequences of GADTs Representations

### 2021-02-14 20-gadts/GADT-story.tex
work on gadt-story Sec 2.2

{\color{blue} polynomial} GADTs relax the restriction on the type
instances appearing in a data type definition. The special form of
GADTs known as {\em nested types} allow the data constructors of a
GADT to take as arguments data whose types involve type instances of
the GADT other than the one being defined. For example, the 
definition 
\[\mathsf{PTree \, a = PLeaf \,a \,|\, PNode \,(PTree\,(a \times
  a))}\] of the nested type $\mathsf{PTree\,a}$ of perfect trees
introduces the data constructor(s $\mathsf{PLeaf :: a \to PTree\,a}$
and) $\mathsf{PNode :: PTree \,(a\times a) \to PTree\,a}$. It enforces
not only the invariant that all of the data in a structure of type
$\mathsf{PTree\,a}$ is of the same type $\mathsf{a}$, but also the
invariant that all perfect trees have lengths that are powers of
2. More general GADTs allow their constructors both to take as
arguments {\em and return as results} data whose types involve type
instances of the GADT other than the one being defined. An example is
the following GADT, given in Haskell syntax:
\[\begin{array}{l}
\mathsf{data\, Seq : * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Const :\, a \to Seq\,a}\\
\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, :\, Seq \,a \to Seq\,b \to
  Seq\,(a \times b)}\\ 
\end{array}\]
Since the return type of the data constructor $\mathsf{Pair}$ is not
simply a variable, $\mathsf{Seq}$ is not a nested type.

@@

To determine the possible shapes of a GADT's container we have to
pattern-match on the type of the data to be contained. For this, our
GADT calculus must support an equality type $\mathsf{Eq}$ that is
inhabited when its two type arguments are the same and is uninhabited
otherwise, i.e., the syntactic reflection of semantic equality. Such a
type can either be taken as primitive or
%, in a higher-kinded calculus such as $F\omega$,
can be represented by its own Church encoding
\[\mathsf{Eq\,t_1\,t_2} = \mathsf{\forall f.\, (\forall a.\,
  f\,a\,a) \to f\,t_1\,t_2}\] 
Using $\mathsf{Eq}$ the GADT $\mathsf{Seq}$, e.g., can be written as 
{\color{red} START HERE}
\[ {\color{blue} WHAT GOES HERE??} \]

Note that Seq has no syntactic map (because Pair doesn't), i.e., is
not an instance of the Functor class, and is not required to in this
representation.

>>>>

{\color{blue} polynomial} GADTs relax the restriction on the type
instances appearing in a data type definition. The special form of
GADTs known as {\em nested types} allow the data constructors of a
GADT to take as arguments data whose types involve type instances of
the GADT other than the one being defined. For example, the 
definition 
\[\mathsf{PTree \, a = PLeaf \,a \,|\, PNode \,(PTree\,(a \times
  a))}\] of the nested type $\mathsf{PTree\,a}$ of perfect trees
introduces the data constructor(s $\mathsf{PLeaf :: a \to PTree\,a}$
and) $\mathsf{PNode :: PTree \,(a\times a) \to PTree\,a}$. It enforces
not only the invariant that all of the data in a structure of type
$\mathsf{PTree\,a}$ is of the same type $\mathsf{a}$, but also the
invariant that all perfect trees have lengths that are powers of
2. More general GADTs allow their constructors both to take as
arguments {\em and return as results} data whose types involve type
instances of the GADT other than the one being defined. An example is
the following GADT, given in Haskell syntax:
\[\begin{array}{l}
\mathsf{data\, Seq :: * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Seq \,a \to Seq\,b \to
  Seq\,(a \times b)}\\ 
\end{array}\]
Since the return type of the data constructor $\mathsf{Pair}$ is not
simply a variable, $\mathsf{Seq}$ is not a nested type.

@@

To determine the possible shapes of a GADT's container we have to
pattern-match on the type of the data to be contained. For this, our
GADT calculus must support an equality type $\mathsf{Eq}$ that is
inhabited when its two type arguments are the same and is uninhabited
otherwise, i.e., the syntactic reflection of semantic equality. Such a
type can either be taken as primitive, as in
\[\begin{array}{l}
\mathsf{data\, Eq :: * \to * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Refl ::\, Eq\,a\,a}\\
\end{array}\]
\noindent
or
%, in a higher-kinded calculus such as $F\omega$,
represented by its own Church encoding
\[\mathsf{Eq\,t_1\,t_2} = \mathsf{\forall f.\, (\forall a.\,
  f\,a\,a) \to f\,t_1\,t_2}\] 
Using $\mathsf{Eq}$, the GADT $\mathsf{Seq}$, e.g., can be written as 
{\color{red} START HERE}
\[ {\color{blue} WHAT GOES HERE??} \]

Note that neither Eq nor Seq has a syntactic map (because one for Eq
leads to a contradition as in Daniel's exampe below, and because one
for Seq can't exist because of Pair), i.e., is not an instance of the
Functor class, and is not required to in this representation.

data Equal a b where
  Refl :: Equal a a

eqElim :: Equal a b -> a -> b
eqElim Refl x = x

emptyFromMapEq :: (forall a b c. (a -> b) -> Equal a c -> Equal b c) -> Void
emptyFromMapEq fmap = eqElim (fmap absurd Refl) ()
